Lesson 3
========================================================

***

### What to Do First?

- Opportunity to let the data surprise you
- Features of the data that might lead to unexpected results
- Understand the variables that are most central in the analysis

***

### Pseudo-Facebook User Data
Loading Facebook data
```{r Pseudo-Facebook User Data}
getwd()
setwd("~/GitHub/2015_Data_Analyst_Project_4/lesson3")
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
# install.packages('ggplot2')
library(ggplot2)
names(pf)
qplot(x=dob_day, data=pf) +
  scale_x_discrete(breaks=1:31)
```

***

#### What are some things that you notice about this histogram?
Response:
Large maximum at 1, almost equally distributed otherwise - probably many 1 also means unspecified day of month?

***

### Moira's Investigation
Notes:
Do peoples perception of the audience on facebook match reality? How many people are actually seeing facebook posts?
[Bernstein, M. S., Bakshy, E., Burke, M., & Karrer, B. (2013). Quantifying the invisible audience in social networks. In Proceedings of the SIGCHI Conference on Human Factors in Computing Systems (CHI 2013), pp. 21-30.](http://hci.stanford.edu/publications/2013/invisibleaudience/invisibleaudience.pdf)

***

### Estimating Your Audience Size
Notes:

- Pointed people to a post
- Compared it to the number of people who have actually seen the post for about 1s

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response: new title picture with nature scene

#### How many of your friends do you think saw that post?
Response: 10%

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response: 10%

***

### Perceived Audience Size

Notes:
Many more people viewed the post than was guessed by the users. Actual size of audience was about 4 times as big.

***
### Faceting
Notes:

- Break histogram into twelve histograms with ```facet_wrap(~variable, ncol=...)```
- Similar: ```facet_grid(vertical_var~horizontal_var)```

```{r Faceting}
qplot(x=dob_day, data=pf) +
  scale_x_discrete(breaks=1:31) +
  facet_wrap(~dob_month, ncol=3)
```

#### Let's take another look at our plot. What stands out to you here?
Response:
first of January occurs way more often

***

### Be Skeptical - Outliers and Anomalies
Notes:

- detecting and dealing with outliers is important
- many causes for outliers (accurate data about an extreme case, bad data / limitations of the data  for a normal or extreme case)

***

### Moira's Outlier
Notes:
In Moira's experiment, a few people guessed a number which was way too high (many millions of people), many people guessed a number way too low (no person at all)

#### Which case do you think applies to Moira's outlier?
Response:
- bad data about an extreme case
- not: bad data about a non-extreme case
- not: good data about an extreme case

***

### Friend Count
#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
qplot(x=friend_count, data=pf)
```

#### How is this plot similar to Moira's first plot?
Response:

- long-tailed data: very few people with a huge number of friends (corresponding to very few people guessing a very high facebook post popularity)
- can be common for user-related data

***

### Limiting the Axes
Notes:

```{r Limiting the Axes}
qplot(x=friend_count, data=pf, xlim=c(0,1000))
qplot(x=friend_count, data=pf) +
  scale_x_continuous(limits=c(0,1000))
```

### Exploring with Bin Width

Notes:

In Moira's experiment, a bin width of 1 shows, that regular numbers in multiples of 10 were most common

***

### Faceting Friend Count

```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50))
```

***

### Friends by Gender
Notes:
there seem to be more males with fewer friends and more females with many friends

```{r Friends \'by\' Gender}
qplot(x=friend_count, data=pf, binwidth=25)+
  scale_x_continuous(limits=c(0,1000),breaks=seq(0,1000,50))+
  facet_wrap(~gender,ncol=1)
```

***

### Omitting NA Values
Notes:

- ```subset(data, !is.na(column))``` removes observation where column is missing
- ```na.omit(data)``` removes all observation where any value is missing

```{r Omitting NA Values}
qplot(x=friend_count, data=na.omit(pf), binwidth=25)+
  scale_x_continuous(limits=c(0,1000),breaks=seq(0,1000,50))+
  facet_wrap(~gender,ncol=1)
```

***

### Statistics 'by' Gender
Notes:

- ```table(...)``` tabulates instances of variable
- ```by(var1, var2, summary)``` shows stats for variable var1 depending on the instances of var2

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

***

#### Who on average has more friends: men or women?
Response: women

#### What's the difference between the median friend count for women and men?
Response: 22

#### Why would the median be a better measure than the mean?
Response: because of outliers strongly pulling the median to one direction or the other

***

### Tenure
Notes:
```{r Tenure}
qplot(x=tenure, data=pf, binwidth=30, color=I('black'), fill=I('#099dd9'))
```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
qplot(x=tenure/365, data=pf, binwidth=.25, color=I('black'), fill=I('#099dd9')) + 
  scale_x_continuous(breaks=seq(1,7,1), limits=c(0,7))
```

***

### Labeling Plots
Notes:

- labeling happens with ```xlab``` and ```ylab```
- can work as comments in code also

```{r Labeling Plots}
qplot(x=tenure/365, data=pf, binwidth=.25, color=I('black'), fill=I('#099dd9'),
      xlab="Number of years in Facebook",
      ylab="Number of users in sample") + 
  scale_x_continuous(breaks=seq(1,7,1), limits=c(0,7))
```

***

### User Ages
Notes: creating a plot with age distribution on facebook

```{r User Ages}
qplot(x=age, data=pf, binwidth=5, color=I('black'), fill=I('#099dd9'),
      xlab="Age of Facebook users in years",
      ylab="Number of users in sample") + 
  scale_x_continuous(breaks=seq(0,125,5), limits=c(0,125))
```

#### What do you notice?
Response:

- many users around the age of 20-25
- a big group of users older than 100 - fake?

***

### The Spread of Memes / Lada's Money Bag Meme
Notes:

- memes = ideas that spread through social media
- some memes are time-contextual but do still re-occur

***

### Transforming Data
Notes:

- exponential / order of magnitude growth and decay much better by log scale (binary, natural or 10-based log)
- helps for making data more normally-distributed looking
- be careful if your data contains zeros (log of zero is undefined) - add 1 if possible
- "data is overdispersed"

***

### Add a Scaling Layer
Notes:

- library ```gridExtra``` lets you put many plots into one graph using ```grid.arrange(...)```
- notice how log10 transformation looks much more like a normal distribution
- log transformed data works well for long-tailed variables!
- you can use ```ggplot(aes(x=...), data=...) + geom_histogram()``` corresponding to ```qplot(x=..., data=...)```
- ```ggplot(...)``` can have additional layers like ```scale_x_log10()``` or ```scale_x_sqrt()``` - however, they do not re-label the axis

```{r Add a Scaling Layer}
# install.packages('gridExtra')
library(gridExtra)
p1 <- qplot(x=friend_count, data=pf)
p2 <- qplot(x=log10(friend_count+1), data=pf)
p3 <- qplot(x=sqrt(friend_count), data=pf)
grid.arrange(p1,p2,p3, ncol=1)
# or
p1 <- ggplot(aes(x=friend_count), data=pf) + geom_histogram()
p2 <- p1 + scale_x_log10()
p3 <- p1 + scale_x_sqrt()
grid.arrange(p1,p2,p3, ncol=1)
```

***


### Frequency Polygons
Notes:

- use ```qplots``` ```geom='freqplot'``` option and setting ```color=...``` to a variable lets you draw frequency polygons by different facets
- good for comparing distributions
```{r Frequency Polygons}
qplot(x=friend_count,
      data=subset(pf, !is.na(gender)),
      binwidth=10, geom='freqpoly', color=gender) +
  scale_x_continuous(lim=c(0,1000), breaks=seq(0,1000,50))
qplot(x=friend_count, y = ..count../sum(..count..),
      data=subset(pf, !is.na(gender)),
      xlab='Friend Count',
      ylab='Proportion of Users with that friend count',
      binwidth=10, geom='freqpoly', color=gender) +
  scale_x_continuous(lim=c(0,1000), breaks=seq(0,1000,50))
```

***

### Likes on the Web
```{r Likes on the Web}
qplot(x=www_likes, data=subset(pf, !is.na(gender)),
      y = ..count../sum(..count..),
      xlab='Number of likes on the web',
      ylab='Proportion of Users with this number of likes on the web',
      geom='freqpoly', color=gender) +
  scale_x_log10()
by(pf$www_likes, pf$gender, sum)
```

***

### Are Likes more often used by males or females
Answer: females

Notes: use ```by(..., sum)``` to facet the data and get a sum

```{r Likes on the Web for male and female}
by(pf$www_likes, pf$gender, sum)
```


***

### Box Plots
Notes: use ```qplot(x=..., y=..., geom='boxplot')``` to plot boxplots with x being the categorical variable and y the variable to examine

```{r Box Plots}
qplot(y=friend_count, x=gender,
      data=subset(pf, !is.na(gender)),
      geom='boxplot')
```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.
Notes:

- ```qplot(...) + scale_y_continuous(limit=c(...))```
- ```qplot(..., ylim=c(...))```
- ```qplot(...) + coord_cartesian(ylim=c(...))``` (better, does remove not remove data points)

```{r}
qplot(y=friend_count, x=gender,
      data=subset(pf, !is.na(gender)),
      geom='boxplot',
      ylim=c(0,1000))
# better:
qplot(y=friend_count, x=gender,
      data=subset(pf, !is.na(gender)),
      geom='boxplot') + 
  coord_cartesian(ylim=c(0,1000))
```

***

### Box Plots, Quartiles, and Friendships
Notes:

- black line: median
- box represents middle 50% of the data

```{r Box Plots, Quartiles, and Friendships}
qplot(x=gender, y=friend_count, data=subset(pf, !is.na(gender)), geom='boxplot') +
  coord_cartesian(ylim=c(0,250))
```

#### On average, who initiated more friendships in our sample: men or women?
Response: female

#### Write about some ways that you can verify your answer: 
Response: use the "by" function to facet friend_count by gender and get the mean: ```by(pf$friend_count, pf$gender, mean)```

```{r Friend Requests by Gender}
by(pf$friend_count, pf$gender, mean)
```

***

### Getting Logical
Notes:

```{r Getting Logical}
summary(pf$mobile_likes)
summary(pf$mobile_likes>0)
pf$mobile_check_in <- NA
pf$mobile_check_in <- ifelse(pf$mobile_likes>0, 1, 0)
pf_mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)
sum(pf$mobile_check_in==1)/length(pf$mobile_check_in)
```

Response:

***

### Analyzing One Variable
Reflection:

- what variables exist? do they have outliers or missing data?
- analyze the data using boxplots, histograms and frequency polygons
- use different scaling (log or sqrt scale) for long-tailed data
- facet data accordingly (facet wraps and multiple layers)
- analyze interquartile ranges, mean, median
- create proxy variables (with logical expressions)
- style up diagrams

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!