---
title: "Problem Set 3"
author: "Benjamin Söllner"
date: "29. September 2015"
output: html_document
---

# 1- Diamonds
```{r}
library(ggplot2)
data(diamonds) # load data
summary(diamonds) # summary of variables
dim(diamonds) # dimensions (observations times variables)
nrow(diamonds) # number of observations
```

**How many observations are in the data set?**: 53940

**How many variables are in the data set?**: 10

**How many ordered factors are in the data set?**: 3

**What represents the best colour?**: D

# 2- Price Histogram

```{r}
qplot(diamonds$price)
```

# 3- Price Histogram Summary
```{r}
summary(diamonds$price)
```

The distribution is skewed to the left and long-tailed. The mean is at 3933, the median at 2401. Further stats:

- Min: 326
- 1st Qu.: 950
- 3rd Qu.: 5324
- Max: 18820

# 4- Diamond Counts
```{r}
nrow(subset(diamonds, price<500))
nrow(subset(diamonds, price<250))
nrow(subset(diamonds, price>=15000))
```

**How many diamonds cost less than $500?**: 1729

**How many diamonds cost less than $250?**: 0

**How many diamonds cost $15,000 or more?**: 1656

# 5- Cheaper Diamonds
Exploring the median region:
```{r}
qplot(diamonds$price, binwidth = 100) +
  scale_x_continuous(limits = c(1500, 3500),
                     breaks = seq(1500, 3500, 100))
```
There are no diamonds that cost $1500.

# 6- Price by Cut Histogram
```{r}
qplot(x=price, data=diamonds, binwidth = 500) +
  facet_wrap(~cut,ncol=3)
```
Similar distributions.

# 7- Price by Cut
```{r}
by(diamonds$price, diamonds$cut, summary, digits = max(getOption('digits')))
```

**Which cut has the highest priced diamond?:** Premium ($18823.00)

**Which cut has the lowest priced diamond?:** Premium & Ideal ($326.00)

**WHich cut has the lowest median price?:** Ideal ($1810.000)

# 8- Scales and Multiple Histograms
```{r}
qplot(x=price, data=diamonds, binwidth = 500) +
  facet_wrap(~cut,ncol=3,scales="free_y")
```
Shows the uniformity of distribution better by making the y-axis scale on demand.

# 9- Price per Carat by Cut
```{r}
qplot(x=price/carat, data=diamonds, binwidth=200) +
  facet_wrap(~cut,ncol=3,scales="free_y")
qplot(x=log10(price/carat), data=diamonds) +
  facet_wrap(~cut,ncol=3,scales="free_y")
```
First showing skewed, second normalized distribution (using log transformation).

# 10- Price Box Plot
```{r}
qplot(y=price, x=cut, data=diamonds, geom='boxplot')
qplot(y=price, x=clarity, data=diamonds, geom='boxplot')
qplot(y=price, x=color, data=diamonds, geom='boxplot')
by(diamonds$price, diamonds$cut, summary)
by(diamonds$price, diamonds$clarity, summary)
by(diamonds$price, diamonds$color, summary)
```

# 11- Interquartile Range - IQR
```{r}
quantile(subset(diamonds, color=="D")$price)
quantile(subset(diamonds, color=="J")$price)
IQR(subset(diamonds, color=="D")$price)
IQR(subset(diamonds, color=="J")$price)
```

**Quartiles for color D**: first: $911.0, third: $4213.5

**Quartiles for color J**: first: $1860.0, third: $7695.0

**IQR for best color (J)**: $3302.5

**IQR for worst color (D)**: $5834.5

# 12- Price per Carat Box Plots by Color
Price per carat of diamonds across the different colors of diamonds using boxplots:

```{r}
qplot(y=price/carat, x=color, data=diamonds, geom='boxplot')
```

# 12- Carat Frequency Polygon 
Investigating the weight of the diamonds (carat) using a frequency polygon.
```{r}
qplot(diamonds$carat, binwidth=0.1, geom='freqpoly') + 
  scale_x_continuous(limits = c(0, 5),
                     breaks = seq(0, 5, 0.2))
```

**Answer:** Only 0.3 and 1.01 have a count greater 2000, 0.1, 0.8, 1.6, 2.0, 3.0 and 5.0 do not!

# 13- Data Wrangling With R

Useful Links: 

- [Data Wrangling in R Intro](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/DataWranglingWithR.pdf)
- [Data Wrangling in R Cheat Sheet](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
- [Other RStudio Cheat Sheets](http://www.rstudio.com/resources/cheatsheets/)

```{r}
# install.packages("tidyr")
# install.packages("dplyr")
```

# 14- Gapminder Data

Analyzing countries by gender about global health: let's see how the global health of different genders is doing in different countries over the years.

```{r fig.width=15, fig.height=20}
library(tidyr)
library(dplyr)
library(ggplot2)

# read data
setwd('h:/Dokumente/GitHub/2015_Data_Analyst_Project_4/lesson3/')
f <- read.csv('Suicide female indicator file UL 20100916.csv', header=T, check.names=F)
m <- read.csv('Suicide male indicator file UL 20100916.csv', header=T, check.names=F)

# rename first column to "country"
colnames(f)[1] <- "country"
colnames(m)[1] <- "country"

# gather year rows into key-value pairs for columns "year" and "fraction"
f.gathered <- tidyr::gather(f, "year", "fraction", 2:length(f))
m.gathered <- tidyr::gather(m, "year", "fraction", 2:length(m))

# add a "gender" column
f.mutated <- dplyr::mutate(f.gathered, gender = "female")
m.mutated <- dplyr::mutate(m.gathered, gender = "male")

# throw data in same pot
data.untidy <- dplyr::bind_rows(f.mutated, m.mutated)

# final cleaning (re-ordering columns, filtering out NAs)
data.tidy <- dplyr::filter(data.untidy[,c(1,2,4,3)], !is.na(fraction))

# Convert year to numeric
data.tidy$year = as.numeric(levels(data.tidy$year))[data.tidy$year]

# Compute most recent years stats
data.now <- dplyr::summarise(dplyr::group_by(data.tidy, country, gender), nowyear=max(year)) %>%
  dplyr::left_join(data.tidy, c("nowyear" = "year", "country" = "country", "gender" = "gender")) %>%
  dplyr::rename(nowfraction=fraction)
# Add summary stats to table
data.tidy <- dplyr::left_join(data.tidy, data.now, c("country" = "country", "gender" = "gender"))


# Graph

# Graph "fraction"" as y and "country" re-ordered by mean.fraction as x
ggplot(data = data.tidy, aes(x=reorder(country, nowfraction), y=fraction)) +
  # Yearly data should appear as a color-coded data point (we are using a reversed terrain colour palette) 
  geom_point(aes(colour=year, alpha=0.5)) +
  scale_colour_gradientn(colours=rev(terrain.colors(length(f)-1)), name="Year") + 
  # We also will plot the mean as a circle
  geom_point(aes(y=nowfraction, shape="O")) + 
  scale_shape_discrete(name="Most Recent", labels=c("Most Recent")) +
  # Facet wrap by gender
  facet_wrap(~gender, ncol=2) + 
  # Themes, Labels, Titles
  coord_flip() + 
  theme_bw() + 
  #theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  xlab("Country") +
  ylab("Suicide Rate per 1000 Capita") +
  ggtitle("Suicide Rate for Different Countries over the Years, separated by Gender")

```

# 15- Analyzing Friends Birthdays
Analyzing distribution of my friend's birthdays

```{r results='hide'}
setwd('h:/Dokumente/GitHub/2015_Data_Analyst_Project_4/lesson3/')
# Loading data
birthdays <- read.csv('friends_birthdays.nogithub.csv', header=T, check.names=F)
# Renaming Columns
colnames(birthdays)[1] <- "Name"
colnames(birthdays)[2] <- "Date"
# Selecting relevant columns
birthdays <- dplyr::select(birthdays, Name, Date)
# Converting to Date
birthdays$Date <- as.Date(birthdays$Date, format="%d.%m.%Y")
# Cleaning: Replacing " birthday" / " Geburtstag" in Label (redundant)
dplyr::transmute(birthdays, Name=gsub(" Geburtstag", "", Name))
# Month and day
birthdays$Month <- as.factor(format(birthdays$Date, "%m"))
birthdays$Day   <- as.factor(format(birthdays$Date, "%d"))
```

## How many people share your birthday? Do you know them?
```{r}
subset(birthdays, Day=="02" & Month=="11")
```

Yes, one is me, the other one I know too.

## How many birthdays are in each month?
```{r}
birthday.months <- dplyr::summarise(dplyr::group_by(birthdays, Month), Count=length(Month))
```

## Which month contains the most number of birthdays?
```{r}
dplyr::arrange(birthday.months, desc(Count))
qplot(data=birthday.months, x=Month, y=Count, geom="bar", stat="identity", fill=I('blue'))
```

August has the most number of birthdays.
