Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes:

- Scatterplots can visualize relationship between two continuous variables
- Perfect accuracy (diagonal line)

***

### Scatterplots
Notes:

- qplot with two continuous variables (x/y) creates scatterplot

```{r Scatterplots}
library(ggplot2)
setwd('h:/Dokumente/GitHub/2015_Data_Analyst_Project_4/lesson4/')
pf <- read.csv('pseudo_facebook.tsv', sep='\t')
qplot(x=age, y=friend_count, data = pf)
qplot(age, friend_count, data=pf)
```

***

#### What are some things that you notice right away?
Response:

- older people seem to have less friends (or their friend_count is in the lower numbers)
- younger people (teenagers) seem to have the most users with many friends and a wide range of friends
- vertical bars (where users might have lied about their ages / fake accounts?)

***

### ggplot Syntax
Notes:

- needs ```geom_...()``` function
- x/y wrapped in ```aes(...)``` wrapper

```{r ggplot Syntax}
ggplot(aes(x=age, y=friend_count), data=pf) + geom_point() +
  xlim(13,90)
summary(pf$age)
```

***

### Overplotting
Notes:

- too man plot points on the same continuous values
- use ```geom_jitter(...)``` to add noise and ```alpha=...``` to overlay points over each other

```{r Overplotting}
ggplot(aes(x=age, y=friend_count), data=pf) + geom_jitter(alpha=1/20) +
  xlim(13,90)
```

#### What do you notice in the plot?
Response:

- still clearly visible "bins" for age ranges
- now much more clearly visible: distribution of friend_counts in each age range

***

### Coord_trans()

Look up the documentation for ```coord_trans()``` and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

Notes:

- use ```coord_trans(x=..., y=...)``` to transform scales, e.g. by ```log10``` or ```sqrt```
- be careful with jittering transformed coordinate systems (noise can move data into negative numbers, leads to imaginary results)
- to specify that just x needs to be jittered, supply ```position = position_jitter(h = 0)```

```{r Coord_trans()}
ggplot(aes(x=age, y=friend_count), data=pf) +
  geom_point(alpha=1/20, position = position_jitter(h = 0)) +
  xlim(13,90) +
  coord_trans(y="sqrt")
```

***

### Alpha and Jitter
Notes:
- Explore relationship between friends initiated vs. age
- Similar distribution than friend count but more homogenous distributions in lower ages

```{r Alpha and Jitter}
ggplot(aes(x=age, y=friendships_initiated),  data=pf) +
  geom_point(alpha=1/20, position = position_jitter(h = 0)) +
  xlim(13,90) +
  coord_trans(y="sqrt")
```

***

### Overplotting and Domain Knowledge
Notes:

- It might make sense to see the x/y variable in terms of percentiles of another variable...

***

### Conditional Means
Notes:

- use ```dplyr::filter(...)```, ```dplyr::arrange(...)```, ```dplyr::group_by(...)```, ```dplyr::mutate(...)```, ```dplyr::summarize(...)``` for data munging

```{r Conditional Means}
library(dplyr)
age.groups <- group_by(pf, age)
pf.fc_by_age <- summarise(age.groups,
          friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n())
pf.fc_by_age <- arrange(pf.fc_by_age, age)
head(pf.fc_by_age)

# also works:

pf.fc_by_age <- pf %>%
  group_by(age) %>%
  summarise(friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n()) %>%
  arrange(age)
```

Create your plot!

```{r Conditional Means Plot}
ggplot(aes(x=age, y=friend_count_mean), data=pf.fc_by_age) +
  geom_line() +
  geom_point() + 
  xlim(13,90) +
  coord_trans(y="sqrt")
```

***

### Overlaying Summaries with Raw Data
Notes:

- ```geom_line(stat='summary', fun.y=f)``` layer adds a line with any function ```f``` applied to the y variables (e.g., ```mean```)

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x=age, y=friend_count),  data=pf) +
  geom_point(alpha=1/20, position = position_jitter(h = 0), color='orange') +
  geom_line(stat='summary', fun.y=mean) + 
  geom_line(stat='summary', fun.y=quantile, probs = .1, linetype=2, color='blue') + 
  geom_line(stat='summary', fun.y=quantile, probs = .9, linetype=2, color='blue') + 
  geom_line(stat='summary', fun.y=quantile, probs = .5, color='blue') + 
  coord_cartesian() + 
  xlim(13,90) +
  coord_trans(y="sqrt")
```

#### What are some of your observations of the plot?
Response:

- the mean and median follow the same path over the age groups
- however, in each age group, the distribution is slightly skewed to the left
- there is a high spike in the age of 69 which resembles a teenager age
- over 75 the data goes crazy
- more than 1000 users is very rare

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

- [Bernstein, M. S., Bakshy, E., Burke, M., & Karrer, B. (2013). Quantifying the invisible audience in social networks. In Proceedings of the SIGCHI Conference on Human Factors in Computing Systems (CHI 2013), pp. 21-30.](http://hci.stanford.edu/publications/2013/invisibleaudience/invisibleaudience.pdf)
- Histograms can be used to pair raw data to summary data
- Show: How much people over-/underestimate by a certain percentage range (= histogram bin)

***

### Correlation
Notes:

- Summarizes a strength of a relation between two variables
- Pearson Product Moment Correlation

```{r Correlation}
cor.test(pf$age, pf$friend_count, method='pearson')
with(pf, cor.test(age, friend_count, method='pearson'))
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response: -0.02740737

***

### Correlation on Subsets
Notes:

- subset the data frame in order to throw out outliers / "fake" data (above age 75)
- ```with(var, ...)``` allows to construct an R environment with variable vectors from a data frame

```{r Correlation on Subsets}
with(subset(pf, age<=70), cor.test(age, friend_count))
```

***

### Correlation Methods
Notes:

- [Correlation Methods: Pearson's r, Spearman's rho, and Kendall's tau](http://www.statisticssolutions.com/correlation-pearson-kendall-spearman/)
- Can be used in ```method``` parameter of ```cor.test(...)```

***

## Create Scatterplots
Notes:

- creating scatterplot of ```likes_received``` vs. ```www_likes_received```, limiting outliers
- adding line of best fit with slope being the corellation coefficient (```geom_smooth(method="lm", ...)```)
- In case ```likes_received``` vs. ```www_likes_received``` variables are closely related because one is the subset of the other

```{r}
ggplot(aes(x=www_likes_received, y=likes_received),  data=pf) +
  geom_point(alpha=1/20) +
  coord_cartesian() +
  xlim(0,quantile(pf$www_likes_received, 0.95)) +
  ylim(0,quantile(pf$likes_received, 0.95)) + 
  geom_smooth(method="lm", color="red")
```

***

### Strong Correlations

Notes: 

- In case ```likes_received``` vs. ```www_likes_received``` variables are closely related because one is the subset of the other

```{r Strong Correlations}
cor.test(pf$www_likes_received, pf$likes_received)
```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

Response: 0.9479902

***

### Moira on Correlation
Notes:

- Measuring correlation is important because sometimes regression models assume that variables are independent from each other
- In case of highly correlated (dependent) variables it will be difficult to tell, which variable causes a certain outcome

***

### More Caution with Correlation
Notes:

- Corellation helps us determine which variables are related / dependent on each other
- But careful: Corellation coefficients can be decepted

```{r More Caution With Correlation}
# install.packages('alr3')
library(alr3)
data(Mitchell)
```

Create your plot!

```{r Temp vs Month}
ggplot(aes(x=Temp, y=Month), data=Mitchell) +
  geom_point() +
  coord_cartesian() +
  geom_smooth(method="lm", color="red")
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

Answer: around 0?

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

Answer: 0.05747063

```{r Noisy Scatterplots}
cor.test(Mitchell$Month, Mitchell$Temp)
```

***

### Making Sense of Data
Notes:

- The months keep repeating over and over
- Looking at the data on a year-by-year base with ```scale_x_discrete```

```{r Making Sense of Data}
ggplot(aes(y=Temp, x=Month), data=Mitchell) +
  geom_point() + 
  geom_line() + 
  coord_cartesian() +
  scale_x_discrete(breaks = seq(0,203,12))
```

***

### A New Perspective

What do you notice?

Response: Peaks in every summer!

Watch the solution video and check out the Instructor Notes!

Notes:

- The cor and cor.test functions determine the strength of a linear relationship, but they may miss other relationships in the data

***

### Understanding Noise: Age to Age Months
Notes:

- Working with "age with months" instead of only "age in years" we should get a smoother data plot 
- Age with months is calculated as ```pf$age + (12-pf$dob_month)/12```

```{r Understanding Noise: Age to Age Months}
library(ggplot2)
ggplot(aes(x=age, y=friend_count_mean), data=pf.fc_by_age) + 
    geom_line()
pf$age_with_months <- pf$age + (12-pf$dob_month)/12
```

***

### Age with Months Means

Notes:

- Now we group by age with month and calculate the aggregate mean

```{r Age with Months Means}
library(dplyr)
pf.fc_by_age_months <- pf %>%
  dplyr::group_by(age_with_months) %>%
  dplyr::summarise(friend_count_mean=mean(friend_count),
                 friend_count_median=median(friend_count),
                 n=n()) %>%
  dplyr::arrange(age_with_months)

head(pf.fc_by_age_months)
```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}
library(ggplot2)
ggplot(aes(x=age, y=friend_count_mean), data=subset(pf.fc_by_age, age<71)) + 
  geom_line() + 
  geom_line(aes(x=age_with_months, y=friend_count_mean), data=subset(pf.fc_by_age_months, age_with_months<71), color='red')

```

***

### Smoothing Conditional Means
Notes:

- using "age with month" effectively decreased the bin size for each sample - therefore the noise increases - **Bias/Variance Tradeoff**
- 

```{r Smoothing Conditional Means}
p1 <- ggplot(aes(x=age, y=friend_count_mean), data=subset(pf.fc_by_age, age<71)) + 
  geom_line() + geom_smooth()
p2 <- ggplot(aes(x=age_with_months, y=friend_count_mean), data=subset(pf.fc_by_age_months, age_with_months<71)) + 
  geom_line() + geom_smooth()
p3 <- ggplot(aes(x=round(age/5), y=friend_count_mean), data=subset(pf.fc_by_age, age<71)) + 
  geom_line(stat='summary', fun.y=mean) + geom_smooth()
library(gridExtra)
grid.arrange(p2, p1, p3, ncol=1)
```

***

### Which Plot to Choose?
Notes:

- In exploratory data analysis you do not often have to choose
- You can create different plots. :)
- Later versions of plots are not always better than earlier ones, just show different insights

***

### Analyzing Two Variables

Reflection:

- how to fine-tune bin sizes in R
- review on how to use dplyr
- review on corellation and how to use it in R
- review on regression and how to use it in R
- with(...) function which resolved for me, why in R we sometimes write df$var and sometimes just var
- really understanding the %>% operator
- review on how to plot two variables against each other (already played around with that during the Gapminder data in problem set 3)

Reflection from Instructors:

- scatter plots and conditional summaries in them
- using jitter and transparancy
- get some insights when to really trust our data (and when rather not to)

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

