---
title: "Problem Set 4"
author: "Benjamin Söllner"
date: "2. Oktober 2015"
output: html_document
---

# price vs. x


```{r}
# In this problem set, you'll continue
# to explore the diamonds data set.

# Your first task is to create a
# scatterplot of price vs x.
# using the ggplot syntax.

# This assignment is not graded and
# will be marked as correct when you submit.

# ENTER YOUR CODE BELOW THIS LINE
# ===========================================
data(diamonds)
library(ggplot2)
ggplot(aes(x=x, y=price), data=diamonds) + 
  geom_point(alpha=0.1)
```

# Findings: price vs. x

**Observation:** exponential growth of price with x, some clusters around integer values of x, some outliers.

# Correlation

```{r}
cor.test(diamonds$x, diamonds$price)
cor.test(diamonds$y, diamonds$price)
cor.test(diamonds$z, diamonds$price)
```

**Between price and x:** 0.8844352

**Between price and y:** 0.8654209

**Between price and z:** 0.8612494

# price vs. depth

```{r}
# Create a simple scatter plot of price vs depth.

# This assignment is not graded and
# will be marked as correct when you submit.

# ENTER YOUR CODE BELOW THIS LINE
#==================================================
data(diamonds)
library(ggplot2)
ggplot(aes(x=depth, y=price), data=diamonds) + 
  geom_point(alpha=0.1)
```

# Adjustments - price vs. depth

```{r}
# Change the code to make the transparency of the
# points to be 1/100 of what they are now and mark
# the x-axis every 2 units. See the instructor notes
# for two hints.

# This assignment is not graded and
# will be marked as correct when you submit.

# ALTER THE CODE BELOW THIS LINE
#============================================
data(diamonds)
library(ggplot2)
ggplot(data = diamonds, aes(x = depth, y = price)) +   geom_point(alpha=0.01) +
  scale_x_continuous(breaks = seq(0,80,2))
```

Hint 1: Use the alpha parameter in ```geom_point()``` to adjust the transparency of points. 

Hint 2: Use ```scale\_x\_continuous()``` with the ```breaks``` parameter to adjust the x-axis.

# Typical Depth Range

**Most diamonds are between which values?** 60 - 64

# Corellation - depth vs. price

```{r}
data(diamonds)
library(ggplot2)
cor.test(diamonds$depth, diamonds$price)
```

**What's the corellation depth vs. price?:** -0.0106474

**Use it to predict the price of the diamond?:** No

**Why?:** No linear correlation - actually increase in variability around 62. 

# price vs. carat

```{r}
# Create a scatterplot of price vs carat
# and omit the top 1% of price and carat
# values.

# This assignment is not graded and
# will be marked as correct when you submit.

# ENTER YOUR CODE BELOW THIS LINE
# ========================================
data(diamonds)
library(ggplot2)
ggplot(aes(x=carat, y=price), data=diamonds) +
  geom_point(alpha=0.1) + 
  xlim(0,quantile(diamonds$carat, 0.99)) + 
  ylim(0,quantile(diamonds$price, 0.99))
```

# price vs. volume

```{r}
# Create a scatterplot of price vs. volume (x * y * z).
# This is a very rough approximation for a diamond's volume.

# Create a new variable for volume in the diamonds data frame.
# This will be useful in a later exercise.

# Don't make any adjustments to the plot just yet.

# This assignment is not graded and
# will be marked as correct when you submit.

# ENTER YOUR CODE BELOW THIS LINE
# =================================================================
data(diamonds)
library(ggplot2)
diamonds$volume <- diamonds$x * diamonds$y * diamonds$z
ggplot(aes(x=volume, y=price), data=diamonds) +
  geom_point(alpha=0.01) + 
  xlim(0,quantile(diamonds$volume, 0.99)) + 
  ylim(0,quantile(diamonds$price, 0.99))
```

# Findings: price vs. volume

**Observations?:**

- most values fall in the range of a lower volume
- increased variability in price with higher volume
- generally increased price with increased volume
- some clustering of volumes occurs
- outliers: some volumes are 0!

# Corellations on Subsets

```{r}
data(diamonds)
library(ggplot2)
diamonds$volume <- diamonds$x * diamonds$y * diamonds$z
diamonds.subset <- subset(diamonds, volume < 800 & volume >= 0)
cor.test(diamonds.subset$price, diamonds.subset$volume)
```

**Corellation price vs. volume when diamonds with volume 0 or >=800 are excluded?:** 0.9216132

Hint 1: It’s helpful to add a variable for volume to the diamonds data frame. You should exclude diamonds that have a volume greater than or equal to 800. Also, exclude diamonds that have a volume of 0. Use the & symbol between two conditions when you subset the original data frame. 

Hint 2: Subset the diamonds data frame under the conditions and save it into another variable. Use that data frame in the cor() function.

# Adjustments - price vs. volume

```{r}
# Subset the data to exclude diamonds with a volume
# greater than or equal to 800. Also, exclude diamonds
# with a volume of 0. Adjust the transparency of the
# points and add a linear model to the plot. (See the
# Instructor Notes or look up the documentation of
# geom_smooth() for more details about smoothers.)

# We encourage you to think about this next question and
# to post your thoughts in the discussion section.

# Do you think this would be a useful model to estimate
# the price of diamonds? Why or why not?

# This assignment is not graded and
# will be marked as correct when you submit.

# ENTER YOUR CODE BELOW THIS LINE
# ========================================

# Prepare Data
data(diamonds)
library(ggplot2)
diamonds$volume <- diamonds$x * diamonds$y * diamonds$z
diamonds.subset <- subset(diamonds, volume < 800 & volume >= 0)

# Make plot
ggplot(aes(x=volume, y=price), data=diamonds.subset) +
  geom_point(alpha=0.01) + 
  xlim(0,800) + 
  geom_smooth(method="lm", color="blue", size=1)

```

This would not neccessary be a good model since the variability of the price increases with increasing volume.

# Mean Price by Clarity

```{r}
# Use the function dplyr package
# to create a new data frame containing
# info on diamonds by clarity.

# Name the data frame diamondsByClarity

# The data frame should contain the following
# variables in this order.

#       (1) mean_price
#       (2) median_price
#       (3) min_price
#       (4) max_price
#       (5) n

# where n is the number of diamonds in each
# level of clarity.

# This assignment WILL BE automatically
# graded!

# DO NOT ALTER THE NEXT THREE LINES OF CODE.
# ======================================================
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
data(diamonds)

# ENTER YOUR CODE BELOW THIS LINE
# ======================================================
data(diamonds)
diamondsByClarity <- diamonds %>%
  dplyr::group_by(clarity) %>% 
  dplyr::summarise(mean_price = mean(price),
        median_price = median(price),
        min_price = min(price), 
        max_price = max(price),
        n = n())
```

# Bar Charts of Mean Price

```{r}
# We’ve created summary data frames with the mean price
# by clarity and color. You can run the code in R to
# verify what data is in the variables diamonds_mp_by_clarity
# and diamonds_mp_by_color.

# Your task is to write additional code to create two bar plots
# on one output image using the grid.arrange() function from the package
# gridExtra.

# This assignment is not graded and
# will be marked as correct when you submit.

# See the Instructor Notes for more info on bar charts
# and for a hint on this task.

# DO NOT DELETE THE LINES OF CODE BELOW
# ===================================================================
data(diamonds)
suppressMessages(library(gridExtra))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))

diamonds_by_clarity <- group_by(diamonds, clarity)
diamonds_mp_by_clarity <- summarise(diamonds_by_clarity, mean_price = mean(price))

diamonds_by_color <- group_by(diamonds, color)
diamonds_mp_by_color <- summarise(diamonds_by_color, mean_price = mean(price))

# ENTER YOUR CODE BELOW THIS LINE
# ===================================================================
p1 <- ggplot(data=diamonds_mp_by_clarity, aes(x=clarity, y=mean_price)) + geom_bar(stat="identity")
p2 <- ggplot(data=diamonds_mp_by_color, aes(x=color, y=mean_price)) + geom_bar(stat="identity")
grid.arrange(p1, p2, ncol=2)
```

# Trends in Mean Price

**Observations in each of the bar charts for mean price by clarity and mean price by color?**

- increase in price for colours E->J
- sharp maximum at clarity SI2 and sharp minimum at VVS1 - strange :( - I would assume that price increases as clarity increases

# Gapminder Revisited

```{r}
# The Gapminder website contains over 500 data sets with information about
# the world's population. Your task is to continue the investigation you did at the
# end of Problem Set 3 or you can start fresh and choose a different
# data set from Gapminder.

# If you’re feeling adventurous or want to try some data munging see if you can
# find a data set or scrape one from the web.

# In your investigation, examine pairs of variable and create 2-5 plots that make
# use of the techniques from Lesson 4.

# You can find a link to the Gapminder website in the Instructor Notes.

# Once you've completed your investigation, create a post in the discussions that includes:
#       1. the variable(s) you investigated, your observations, and any summary statistics
#       2. snippets of code that created the plots
#       3. links to the images of your plots

# Copy and paste all of the code that you used for
# your investigation, and submit it when you are ready.
# ====================================================================
suppressMessages(library(dplyr))
suppressMessages(library(tidyr))

setwd('h:/Dokumente/GitHub/2015_Data_Analyst_Project_4/lesson3/')
mentalhealth <- read.csv('Suicide male indicator file UL 20100916.csv', header=T, check.names=F)
mentalhealth <- mentalhealth %>%
  tidyr::gather("year", "fraction", 2:length(mentalhealth)) %>%
  dplyr::filter(!is.na(fraction))
mentalhealth$year = as.numeric(levels(mentalhealth$year))[mentalhealth$year]
colnames(mentalhealth)[1] <- "country"
mentalhealth.now <- mentalhealth %>%
  dplyr::group_by(country) %>% 
  dplyr::summarise(nowyear=max(year)) %>%
  dplyr::left_join(mentalhealth, c("nowyear" = "year", "country" = "country")) %>%
  dplyr::rename(nowfraction=fraction)
```

Material: 

- [Strong Relationship beteen Income and Mental Health by Frank Newport](http://www.gallup.com/poll/102883/strong-relationship-between-income-mental-health.aspx)
- [The Prevalence, Distribution, and Mental Health Correlates of Perceived Discrimination in the United States by Roland C. Kessler et al.](http://www.aging.wisc.edu/midus/findings/pdfs/52.pdf)
- [Mental Health And Criminal Justice by Paul E Mullen et al.](http://crg.aic.gov.au/reports/mullen.pdf)

[Gapminder's Data](http://www.gapminder.org/data/)

- [Inequality index (Gini)](http://spreadsheets.google.com/pub?key=pyj6tScZqmEcjeKHnZq6RIg&output=xls)
- [Income share of the highest 10%](http://spreadsheets.google.com/pub?key=tmKvydPl_roGIQBrMYA6C4g&output=xls)
- [Income share of the lowest 10%](http://spreadsheets.google.com/pub?key=trzLWJQU4SZMDpeVg3XnL5A&output=xls)
- [Extreme Poverty < 125$ per day of 20120905](http://spreadsheets.google.com/pub?key=0ArfEDsV3bBwCdDhjcXdjbURLMFFVcVFPYThhYmtvZXc&output=xls)
- [Government Health Spending](http://spreadsheets.google.com/pub?key=tBwBBkViOJoycBhLnWHqwSQ&output=xls)
  
[Hans Rosling's 200 Countries, 200 Years, 4 Minutes](https://www.youtube.com/watch?v=jbkSRLYSojo)

[Save plots using ```ggsave```](http://docs.ggplot2.org/current/ggsave.html)