Lesson 6
========================================================

### Welcome
Notes:

- diamonds have a fascinating history (drove colonialization and many wars, pushed the south african economy)
- here we will build a model to predict diamond prices

```{r}
data(diamonds)
#setwd("h:/Dokumente/GitHub/2015_Data_Analyst_Project_4/lesson6")
library(ggplot2)
```
***

### Scatterplot Review

```{r Scatterplot Review}
ggplot(data=diamonds, aes(x=carat, y=price)) + 
  geom_point(color=I('orange')) + 
  xlim(0,quantile(diamonds$weight, 0.99)) + 
  ylim(0,quantile(diamonds$price, 0.99)) + 
  stat_smooth(method="lm")
```

***

### Price and Carat Relationship
Response:

- The relationship between price and carat is monotonous, non-linear
- Dispersion in the data over carat increases
- Some clusters on certain carat numbers
- linear model might not predict data very well

***

### Frances Gerety
Notes:

- diamonds dataset ships with ggplot2
- collected for >50.000 diamonds from DiamondSc.info
- history of diamonds - TLDR: diamonds for a long time remained an only rarely demanded jewelery - until a marketing campaign in the US started

#### A diamonds is

... forever -- Frances Gerety, DeBeers cartel (slogan of the century)

***

### The Rise of Diamonds
Notes:

- Tricky marketing campaign was started
- Diamonds given out to celebrities, british monarchs persuaded to wear diamonds over other gems
- Two month benchmark: "Isn't a two-month salary a little price to pay for something that lasts forever?"
- Getting engaged in America = getting a diamond

***

### ggpairs Function
Notes:

```{r ggpairs Function}
# install these if necessary
# install.packages('GGally')
# install.packages('scales')
# install.packages('memisc') # summarize regression
# install.packages('lattice')
# install.packages('MASS')
# install.packages('car') # recode variables
# install.packages('reshape') # reshaping / wrangling
# install.packages('plyr') # summarization

# load the ggplot graphics package and the others
library(ggplot2)
library(GGally)
library(scales)
library(memisc)

# sample 10,000 diamonds from the data set
set.seed(20022012)
diamond_samp <- diamonds[sample(1:length(diamonds$price), 10000), ]
ggpairs(diamond_samp, params = c(shape = I('.'), outlier.shape = I('.')))
```

ggpairs plots

- in lower triangle: grouped histograms for qualitative-\* pairs, scatterplots for quantitative-quantitative pairs
- in upper triangle: grouped histograms for qualitative-qualitative pairs (reversed), boxplots for qualitative-quantitative pairs, correlation for quantitative-quantitative pairs

What are some things you notice in the ggpairs output about the relation for variable ```price```?
Response:

- price increases with x, y, z in a less-than-linear way
- price increases with carat in a more-than-linear way
- suggests a corellation weight ~ volume = x*y*z -> suggests using 3rd root of carat as price predictor
- for "depth" there is a sweet spot, where diversion of price is maximal, outside from it, price is in a lower region
- for the categorical variables "clarity", "color", "cut", prices increase with "color", decrease with "clarity", increase with "cut"


***

### The Demand of Diamonds
Notes:

- the diversion in higher carat diamonds may be due to the lower sensitivity of customers shopping for higher-cost (more expensive) diamonds
- because of the smaller demand for bigger diamonds the market for those diamonds is also less competitive
- usually distributions for monetary variables are very skewed and vary in orders of magnitude due to path dependence (the rich getting richer) and multiplicative processes (like inflation)
- hence, log scale might be applicable for certain prices

```{r The Demand of Diamonds}
library(gridExtra)

p1 <- qplot(diamonds$price, fill=I("orange")) + ggtitle('Price')
p2 <- qplot(diamonds$price, fill=I("blue")) + scale_x_log10() + ggtitle('Price (log10)')
grid.arrange(p1,p2)
```

***

### Connecting Demand and Price Distributions
Notes:

There are two modes = sweet spots where supply meets demand - one for "normal" one for "premium" customers.

***

### Scatterplot Transformation

```{r Scatterplot Transformation}
qplot(carat, price, data=diamonds) + 
  scale_y_continuous(trans=log10_trans()) + 
  ggtitle("Price (log10) by Carat")
```

### Create a new function to transform the carat variable

```{r cuberoot transformation}
cuberoot_trans = function() trans_new('cuberoot', transform = function(x) x^(1/3),
                                      inverse = function(x) x^3)
```

- ```trans_new(name, transform=..., inverse=...)``` creates a new function that can be made into a "pointer" using ```function() trans_new(...)```
- ```transform``` and ```inverse``` must be functions themselves like ```function(x) x^(1/3)``` or ```function(x) x^3```

#### Use the cuberoot_trans function
```{r Use cuberoot_trans}
ggplot(aes(carat, price), data = diamonds) + 
  geom_point() + 
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat')
```

***

### Overplotting Revisited

- Overplotting is normal in big datasets
- Can be seen by sorting the variable and looking through its values
- Overplotting can obscure information about density in the data 
- Remedy: smaller points, jittering, alpha ```geom_point(alpha=0.5, size=0.75, position="jitter")```

```{r Sort and Head Tables}
head(sort(table(diamonds$carat), decreasing=T))
head(sort(table(diamonds$price), decreasing=T))
```


```{r Overplotting Revisited}
ggplot(aes(carat, price), data = diamonds) + 
  geom_point(alpha=0.5, size=0.75, position="jitter") + 
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat')
```

***

### Other Qualitative Factors
Notes:

- Clarity also seem to be an influencing factor (albeit not that much) - can mostly be seen by jewelers with a jewelers loop 
- Cut also has a slightly higher impact ("firy quality of diamonds")

***

### Price vs. Carat and Clarity

Alter the code below.
```{r Price vs. Carat and Clarity}
# install and load the RColorBrewer package
#install.packages('RColorBrewer')
library(RColorBrewer)

ggplot(aes(x = carat, y = price, color = clarity), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Clarity', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
    breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
    breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Clarity')
```

***

### Clarity and Price
Response:

Clarity explains quite a lot of the variance for a certain carat value (clearer diamonds are more expensive, unclear diamonds are seen as "imperfect").

***

### Price vs. Carat and Cut

```{r Price vs. Carat and Cut}
ggplot(aes(x = carat, y = price, color = cut), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Cut', reverse = T,
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Cut')
```

***

### Cut and Price
Response:

Not much variance - most diamonds are "ideal" cut anyway.

***

### Price vs. Carat and Color

Alter the code below.
```{r Price vs. Carat and Color}
ggplot(aes(x = carat, y = price, color = color), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = "Color",
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Color')
```

***

### Color and Price
Response:

Yes, but color shouldn't be visible to the naked eye.

***

### Linear Models in R
Notes:

- linear models in R can be built like so: ```lm(outcome_var ~ explanatory_var)```
- for diamonds dataset: ```lm(log(price) ~ carat^(1/3))```

Response:

***

### Building the Linear Model
Notes:

- ```I(...)``` wrapper: "as-is" - keeps R from evaluating an expression so it can be given to a model
- updating / re-fitting models: ```update(model, ...)```
- resulting model: ```log(price) ~ carat^(1/3) + cut + color + clarity```
- that means: ```log(price) = a0 + a1 * carat^(1/3) + a2 * cut + a3 * color + a4 * clarity + epsilon``` (```epsilon``` = residual)
- ```mtable(...)``` analyzes model (we see that we account for 92.4% of the variance)

```{r Building the Linear Model}
m1 <- lm(I(log(price)) ~ I(carat^(1/3)), data = diamonds)
m2 <- update(m1, ~ . + carat)
m3 <- update(m2, ~ . + cut)
m4 <- update(m3, ~ . + color)
m5 <- update(m4, ~ . + clarity)
mtable(m1, m2, m3, m4, m5)
```

Notice how adding cut to our model does not help explain much of the variance
in the price of diamonds. This fits with out exploration earlier.

***

### Model Problems
Video Notes:

- Model: ```ln(price) = 0.415 + 9.144 * carat^(1/3) - 1.093 * carat + ... * cut + ... * color + ... * clarity + epsilon
- What are some problems for this model?

Response:

- Diamond prices might not be consistent over time (2008 -> global recession, prices plummeting, afterwards 6%/year increase, also: Diamond market in China heating up)
- Price increases of the diamonds might even vary for different carat sizes
- Where diamond is sourced may play a role
- The factors might not be truly independent from each other (e.g., cut/color/clarity might be more important for bigger diamonds)

***

### A Bigger, Better Data Set
Notes:

- The code used to obtain the data is available here: [https://github.com/solomonm/diamonds-data]
- Can be downloded and loaded with ```load(...)``` instead

```{r A Bigger, Better Data Set}
#install.package('bitops')
#install.packages('RCurl')
#library('bitops')
#library('RCurl')

#diamondsurl=getBinaryURL('https://github.com/SolomonMg/diamonds-data/blob/master/BigDiamonds.Rda')
#load(rawConnection(diamondsurl))
load("BigDiamonds.Rda")
```


## Building a Model Using the Big Diamonds Data Set
Notes:

We will re-fit the model, this time also account for outliers and only choose diamonds certified by GIA

```{r Building a Model Using the Big Diamonds Data Set}
m1 <- lm(I(log(price)) ~ I(carat^(1/3)), data = diamondsbig)
m2 <- update(m1, ~ . + carat)
m3 <- update(m2, ~ . + cut)
m4 <- update(m3, ~ . + color)
m5 <- update(m4, ~ . + clarity)
mtable(m1, m2, m3, m4, m5)

```

***

## Predictions

Example Diamond from BlueNile:
Round 1.00 Very Good I VS1 $5,601

```{r}
#Be sure you’ve loaded the library memisc and have m5 saved as an object in your workspace.
thisDiamond = data.frame(carat = 1.00, cut = "V.Good",
                         color = "I", clarity="VS1")
modelEstimate = predict(m5, newdata = thisDiamond,
                        interval="prediction", level = .95)
exp(modelEstimate)
```

Evaluate how well the model predicts the BlueNile diamond's price. Think about the fitted point estimate as well as the 95% CI.

- fitted value is 8.473461 with lower CI 8.017327 and upper CI 8.929595
- price has still to be taken to the ```exp(...)``` power: 4,786$ (lower CI 3,033\$, upper CI 7,552\$)

***

## Final Thoughts
Notes:

- it is not irrelevant where you buy diamonds

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

