Lesson 5
========================================================

### Multivariate Data
Notes:

- last lesson: 2 quantitative variables
- now: 3 or more variables

***

### Moira Perceived Audience Size Colored by Age
Notes:

- investigation whether age has an effect in perceived audience size
- added 3rd variable represented with color
- was not very effective

***

### Third Qualitative Variable
Notes:

- sometimes in exploratory data analysis we reach dead end
- does age have an effect on friend count?
- maybe there is a totally different age distribution for male & female -> draw a boxplot (x = ```shape=4```) denotes mean and a line distribution
- also, we create a summary data frame ```pf.fc_by_age_gender```

```{r Third Qualitative Variable}
library(ggplot2)
setwd('h:/Dokumente/GitHub/2015_Data_Analyst_Project_4/lesson4/')
pf <- read.csv('pseudo_facebook.tsv', sep='\t')

ggplot(aes(x=gender, y=age),
       data=subset(pf, !is.na(gender))) +
  geom_boxplot() + 
  stat_summary(fun.y=mean, geom="point", shape=4)

ggplot(aes(x=age, y=friend_count),
       data=subset(pf, !is.na(gender))) + 
  geom_line(aes(color=gender), stat="summary", fun.y=median)

library(dplyr) 
pf.fc_by_age_gender <- pf %>%
  filter(!is.na(gender)) %>%
  group_by(age, gender) %>%
  summarise(mean_friend_count = mean(friend_count),
            median_friend_count = median(friend_count),
            n=n()) %>%
  ungroup() %>%
  arrange(age)
```

***

### Plotting Conditional Summaries
Notes:

- now we can plot the summary data

```{r Plotting Conditional Summaries}
ggplot(aes(x=age, y=median_friend_count),
       data=pf.fc_by_age_gender) + 
  geom_line(aes(color=gender), stat="summary", fun.y=median)
```

***

### Thinking in Ratios
Notes:

- it's sometimes helpful to understand data in terms of fractions
- e.g., how many times more friends does the average female user have at a certain age?

***

### Wide and Long Format
Notes:

- **long format**: age, gender, mean_friend_count etc.
- **wide format**: age, male, female etc.


***

### Reshaping Data
Notes:

- convert between long and wide format with ```gather``` and ```spread``` (```tidyr```` library) or with ```dcast```` and ```melt``` (```reshape2``` library)

```{r}
#install.packages('reshape2')
library(reshape2)
pf.fc_by_age_gender.wide <- dcast(pf.fc_by_age_gender, age ~ gender, value.var = "median_friend_count")
```

***

### Ratio Plot
Notes:

- you can specify the ratio ```female/male``` for ```y```
- adding a horizontal line: ```geom_hline(...)```

```{r Ratio Plot}
library(ggplot2)
ggplot(aes(x=age, y=female/male), data=pf.fc_by_age_gender.wide) + 
  geom_line() + 
  geom_hline(yintercept=1, alpha=0.3, linetype=2)
```

***

### Third Quantitative Variable
Notes:

- how to quantify a difference as ratio when variable is continuous (not categorical)?
- you can make buckets / bins with the ```cut``` function

```{r Third Quantitative Variable}
pf$year_joined =  floor(2014 - pf$tenure/365)
summary(pf$year_joined)
table(pf$year_joined)
```

***

### Cut a Variable
Notes:

- ```cut``` can make discrete variables from continuous ones

```{r Cut a Variable}
pf$year_joined.bucket <- cut(pf$year_joined, c(2004, 2009, 2011, 2012, 2014))
table(pf$year_joined.bucket, useNA="ifany")
```

***

### Plotting it All Together
Notes:

- plotting the dependency on variable ```year_joined.bucket```works very much like plotting the dependency on ```gender```

```{r Plotting it All Together}
library(dplyr)
pf.fc_by_age_year_joined.bucket <- pf %>%
  filter(!is.na(year_joined.bucket)) %>%
  group_by(age, year_joined.bucket) %>%
  summarise(median_friend_count = median(friend_count),
            mean_friend_count = mean(friend_count),
            n=n()) %>%
  ungroup() %>%
  arrange(age)
 
ggplot(data=pf.fc_by_age_year_joined.bucket, 
       aes(x=age, y=median_friend_count)) + 
  geom_line(aes(color=year_joined.bucket)) 

# or:

ggplot(data=subset(pf, !is.na(year_joined.bucket)),
       aes(x=age, y=friend_count)) + 
  geom_line(aes(color=year_joined.bucket),
            stat="summary", fun.y=median)
```

***

### Plot the Grand Mean
Notes:

- Grand Mean can just be plottet by one additional (unspecific col ored) ```geom_line(...)```
- we can see, that much of the data is about recently joined cohorts


```{r Plot the Grand Mean}
ggplot(data=subset(pf, !is.na(year_joined.bucket)),
       aes(x=age, y=friend_count)) + 
  geom_line(aes(color=year_joined.bucket),
            stat="summary", fun.y=mean) + 
  geom_line(stat="summary", fun.y=mean,
            linetype=2)

```

***

### Friending Rate
Notes:

- we can not see much difference in the way friend_count is distributed between different cohorts
- this makes uns confident that the friend_count distribution by age is not dependent on how much time users had to accumulate friends 
- we can use ```friend_count``` and ```tenure``` as a rate (how many friends per time interval does a user aquire?)

```{r Friending Rate}
with(subset(pf, tenure >= 1), summary(friend_count/tenure))
```

**What is the median friend rate?** 0.2205

**What is the maximum friend rate?** 417.0000

***

### Friendships Initiated
Notes:

- do users who are longer on the site initiate more friendships?
- plot friendships initiated vs. tenure

```{r Friendships Initiated}
ggplot(data=subset(pf, !is.na(year_joined.bucket) & tenure >= 1),
       aes(x=tenure, y=friendships_initiated/tenure)) + 
  geom_line(aes(color=year_joined.bucket),
            stat="summary", fun.y=median)
```

- people with more tenure generally initiate less friendships

***

### Bias-Variance Tradeoff Revisited
Notes:

- smoothing out plot with ```aes(x=n*round(var/n), ...)``` increases bin size
- higher bin size -> higher bias -> lower variance / noise
- same effect ```geom_smooth(...)```

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

# Smoother (geom_smoot(...)):

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_smooth(aes(color = year_joined.bucket))
```

***

### Sean's NFL Fan Sentiment Study
Notes:

- how to visualize fan sentiment on sports groups
- ratio of positive to negative words in 5 minute increments over a season of 4 months
- how to get less noisy data?
  - compute moving average (good interval: 1 day) -> still too noisy
  - use a "spline" -> more predictable but too smooth -> doesn't capture jumps in sentiment on game days
  - 7 day average always includes last game -> best depiction of the data
- generally, a bigger window for moving averages increases bias, decreases variance
- combined with spline based model -> best of both worlds (including "jumps" on game day)

***

### Introducing the Yogurt Data Set
Notes:

- data from Danone Yogurt purchases over time by household
- identifier of household is credit card
- representative of typical data mined in retail industry
- structure of Yogurt data is quite different: many rows per household, each row for one purchase occasion
- different kinds of questions may be answered about individual household

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
setwd('h:/Dokumente/GitHub/2015_Data_Analyst_Project_4/lesson5/')
yo <- read.csv('yogurt.csv')
yo$id <- factor(yo$id)
str(yo)

qplot(data=yo, x=price)
```

**What do you notice?**: Increasing price; discreetness (some gaps in between) - can be missed with other binwidth

- ```summary(...)``` does not easily show us the discreetness of the prices
- however, ```length(unique(...))``` can show us, that for a large with of the distribution there are only about 20 different unique values
- ```table(...)``` also works very well (like histograms)

```{r}
summary(yo$price)
length(unique(yo$price))
table(yo$price)
```

***

### Number of Purchases
Notes:

- on a given purchasing occasion, how many Danone yogurts does a household purchase?
- ```transform(...)``` makes it possible to add another variable with the sum of a few variables

```{r Number of Purchases}
yo <- transform(yo, all.purchases = strawberry + blueberry + pina.colada + plain + mixed.berry)
summary(yo$all.purchases)
qplot(yo$all.purchases)
```

We can notice, that most households buy more than one yogurt at a time.

***

### Prices over Time
Notes:

- scatterplot of price vs. time

```{r Prices over Time}
ggplot(data=yo, aes(x=time, y=price)) + 
  geom_point(alpha=0.2)
```

**What do you notice?:**

- overall increase in the price 
- some lower price values (coupons?)

***

### Sampling Observations
Notes:

- when familiarizing with a new dataset that contains multiple observations of the same units, it might be useful to work with samples of the data set (easier to display) 
- yogurt dataset: maybe choose a small number of households in order to assess variability within / between households etc.
- e.g., originally, the yogurt dataset was used to model consumer preference for variety
- questions to be answered: how often do households buy yogurt? how often do they buy multiple items? what prices do they buy yogurt at?
- let's pick 16 households

***

### Looking at Samples of Households

Notes:

- ```set.seed(4230)``` makes the sampling reproducible
- ```sample(levels(factor), n)``` samples ```n``` single samples from a factor
- faceting over ```id``` ("case variable")

```{r Looking at Sample of Households}
set.seed(4230)
sample.ids <- sample(levels(yo$id), 16)

# drawing a little diagram for every household with the number of purchases indicated by the symbol
ggplot(aes(x=time, y=price),
       data=subset(yo, id %in% sample.ids)) + 
  facet_wrap(~id) +
  geom_line() + 
  geom_point(aes(size=all.purchases), pch=1)
```

We could also join in the coupon data to explain lower prices.

***

### The Limits of Cross Sectional Data
Notes:

- cross sectional data (like our facebook data) cannot be faceted by case (only one single row per case)
- if we had better data about cases (users) over time, this could explain better the relation in friendships_initiated dropping over tenure

***

### Many Variables
Notes:

- until now: variable was always pre-chosen
- trying to confirm that relationship between two variables is consistent across third, fourth etc. variable
- but how to identify auxiliary variables? (maybe even 10, 20 or hundreds of others? or summarizing them into a computed subset of relevant variables?)
- there are some tools to create many plots at once

***

### Scatterplot Matrix

Notes:

- grid of scatterplot between every pair of variables with ```ggpairs(...)``` from ```GGally``` library

```{r fig.width=30, fig.height=30}
#install.packages("GGally")
library(GGally)
#theme_set(theme_minimal(20))
set.seed(1836)
pf_subset <- pf[, c(2:15)]
names(pf_subset)
ggpairs(pf_subset[sample.int(nrow(pf_subset), 1000), ])
```

**Correlation between ```friendships_initiated``` and ```friendcount```:** 0.769

**Correlation between ```age``` and ```mobile_likes```:** -0.0475

- the diagonal shows the distribution of every variable
- the plots below the diagonal show plots with x and y maped to either of the variables
- for the variables including "gender" (categorical variable) faceted histograms (for discrete data) or boxplots (for continuous data) are shown
- for some variables, we might want to switch to another scale (like logarithmic) - not possible with ```ggpairs```

***

### Even More Variables
Notes:

- another example: genomic data - extremely many variables (measurements) - over thousands - for even small samples!
- use case: genes that are related to a certain disease
- dataset with expression of 6830 genes related to a tumor with a reference baseline

***

### Heat Maps
Notes:

- here is some data with 64 variables
- first: create a long format (with 64 variables spread out into single rows) using ```melt```
- then: re-naming values and and plotting over-expressed and under-expressed genes by case
- here we only look at 200 genes

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```

***

### Analyzing Three of More Variables
Reflection:

- inferring one variable from others
- reshaping data between long and wide formats
- using smoothers or adding summaries
- using color as another visual cue to display variables
- data with many observation per sample value and with only one observation (over time)
- subsampling data
- creating scatterplot matrices
- creating heatmaps

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

